A Recipe for Scavenging \textcolor{blue}{in Vertebrates} - the natural history of a behaviour

### 2016-07-19 response_letter.tex

\smallskip

\textcolor{blue}{As we note below, we feel our earlier manuscript did not capture the hierarchy we had in mind.This updated version is more explicit in terms of the structure where we now highlight the sections and subsections. Regarding the novelty of our manuscript, we have changed our introduction which was regarded as too similar to deValult et al. (2003) and argue our case for what we consider inevitable overlap with Beasley et al. (2015).}

\section{Reviewer: 1}
This paper attempts to synthesize literature on traits that enable scavenging behavior in extant and extinct vertebrates. The paper is mostly well written, and most (but not all) pertinent literature sources are included. I do worry about the lack of novelty of the manuscript, and what I perceived as haphazard organization. Here are some specific issues that should be addressed in my opinion:

\begin{enumerate}
\item{Abstract: Scavenging can be difficult to observe, but probably no more so than predation. Also, using experimental carcasses, scavenging is actually quite easy to observe.}

\textcolor{blue}{We emphasise that food webs have tended to lump all carnivory into predation and it is with extinct species that we have no observational evidence.}

\item{Abstract: You should point out that your ``scale of scavenging'' is only qualitative; you made no attempt to quantify such traits.}
\item{Introduction: Much of this section seems to follow DeVault's (2003) review paper on scavenging—in fact, there is almost nothing new here that is not covered in that earlier review.}

@@

\item{Introduction: You mention that you will use ``other methods in order to discern the most suitable morphologies...''  but you don't say what they are. What other methods are you talking about?}

\textcolor{blue}{We now specify the approaches we have in mind:  \textit{``e.g. energetics models, comparative anatomy, palaeontology etc.'' }}

\item{I think that early on in the paper you need to define obligate and facultative scavengers and then throughout the paper, specify which you are referring to. For example, in the first sentence of the Detection section, you talk about ``known scavengers'', but it is unclear if you really mean obligate scavengers. Nearly all carnivorous vertebrates are scavengers at least to some extent, so you need to elaborate on what you mean when you talk about ``scavengers''.}

\textcolor{blue} {We make the following point in our introduction: \textit{Indeed, there is no discrete divide between predators and scavengers but rather a continuous gradient. Even vultures, the canonical example of obligate scavengers, can hunt.} We have tried to be clearer throughout the paper as well.}

%TG: maybe say something along the lines:
%\textcolor{blue}{As with the above comment we argue that all carnivorous vertebrates are placed along a gradient of scavenging.}

@@

\smallskip
\textcolor{blue}{One of the central points of our paper is to compare the biology of extant scavengers with extinct species in order to determine if they were suited to scavenging. All of the examples we use are backed up by references from previous work. Rather than taking them as fact we offer these as possible behaviours that the extinct species engaged in. Because we can't observe their behaviour directly we are left to make assumptions about how likely they were to have enganged in them. But this is hardly unique to our work. We have added some extra references in support of our claims: \textit {``Thus, given pterosaurs seem to have cohabited in large numbers (Witton 2013), and the known benefits this can have for social foraging in birds (Dermody et al. 2011), it seems probable that scrounging behaviours were seen in the flying pterosaurs as well.''}}
\smallskip

While the authors distinguish between obligate and facultative scavenging, they tend to lump them together when it comes to constructing their arguments. In the end, is the argument about the former, or the latter? Is the real interest in whether species scavenge at all, what proportion of their resources are obtained by scavenging, or what the conditions are that produce facultative or obligate scavengers?

\textcolor{blue} {This was also raised by reviewer 1. We make the following point in our introduction: \textit{Indeed, there is no discrete divide between predators and scavengers but rather a continuous gradient. Even vultures, the canonical example of obligate scavengers, can hunt.} We have tried to be clearer throughout the paper as well. }

@@

They also seem to ignore flexibility in species' behaviors. For instance, while spotted hyenas are on their highly scavenging side of their Figures 1 and 2, in some ecosystems they hunt significant proportion of their prey. Similarly, lions can take a large proportion of their prey as scavenged, as the authors note. But to suggest, as they do in their abstract, that they can apply their scale of scavenging to any species at any time to judge how important scavenging was in its diet is frankly completely ignoring behavioral variability over time and space. 
\smallskip

\textcolor{blue}{It is certainly true that scavengers can be variable in the proportion of carrion they consume but we would argue that species like hyenas have certain characteristics that make this possible. In contrast, cheetahs do not have a biology that lends itself to scavenging in the wild, due to poor competitive ability etc. We appreciate that there is a continuum across and among species which we specifically note in this revised version. Moreover, researchers working in palaeobiology often try to reconstruct the behaviours of extinct species; we are following this approach when we make inferences about ancient systems.}
\smallskip

@@

It is substantially more difficult with fossil species for which we may know far fewer of the variables they list in Figure 1, in a nuanced way. Figure 2 does not display the diversity of scavengers through time, it displays examples of species that display observed or inferred scavenging behaviors. Diversity has specific ecological definition(s) that are being ignored here.

If the authors chose to resubmit, I would suggest they organize their manuscript based on known information about modern species, known (e.g. bone processing by theropods) and inferred (e.g. scavenging in hominins) information about fossil species, and then try to draw broader conclusions. They might look to articles about other modern scavenging opportunities such as Capaldo and Peter's study of wildebeest drownings \url{http://www.sciencedirect.com/science/article/pii/S0305440385700398}, and suggestions that sabertooth cats which they claim were unable to feed on bones may have consumed more bone than penecontemporaneous dire wolves \url{https://www.jstor.org/stable/20627167?seq=1#page_scan_tab_contents}. \\
\smallskip
\textcolor{blue}{The issue of the structure of our paper was noted by all reviewers who all had suggestions to improve it. 
We have made an effort to delineate all of the main sections and subsections so that our frame of optimal foraging is more apparent.}
\smallskip
\\
In sum, while I think the aim of the article is a very worthwhile one, I think the authors cannot claim to have produced an all-encompassing model that can predict the presence or level of scavenging in any organism - because I think this simply cannot be done.

@@

Finally, some of the perspectives of this manuscript overlap substantially with parts of Beasley et al. 2015 (pages 108-112), which is cited numerous times in the manuscript.
\smallskip

\textcolor{blue}{The sections of Beasley et al. (2015) are:  1. the evolution of scavengers, 2. competition between scavengers and 3. environmental effects on scavenger ecology. These are very broad areas so in a review of this sort some overlap is inevitable. We feel our inclusion of extinct species and systems as well as the frame of optimal foraging theory gives this manuscript its novelty.} 
\smallskip

### 2016-07-19 natural history of scavenging.tex
updates on ms and letter

\begin{abstract} 
Despite its prevalence, scavenging is a difficult behaviour to observe in modern day carnivores and impossible to study directly in extinct species. 
Yet, there are certain intrinsic and environmental features of a species that push it towards a scavenging lifestyle. 
These can be thought of as some of the principal parameters in optimal foraging theory namely, encounter rate, handling time and prey availability. 
We use these components to highlight the morphologies and environments that would have been conducive to scavenging over geological time by focusing on the dominant vertebrate groups of the land, sea and air. 
The result is a synthesis on the natural history of scavenging. 
Our idea of a scale of scavenging can be applied to any species at any time to help \textcolor{blue}{judge} the importance of this behaviour in its diet. 
\end{abstract}

@@

Historically, \textcolor{blue}{food webs have classified most acts of carnivory as predation events \citep{wilson2011scavenging}. 
This is clearly a mistaken view because of the existence of scavenging, a behaviour displayed by almost all carnivorous vertebrates \citep{devault2003scavenging}. 
Recent research has begun to redress this imbalance by recognising the prevalence of scavenging and detailing its ecological consequences across modern ecosystems where we are beginning to appreciate the implications for the trophic ecology of these systems \citep{pereira2014facultative,periquet2015lion}.
Indeed, there is no discrete divide between predators and scavengers but rather a continuous gradient \citep{pereira2014facultative}.
Unfortunately, the lack of direct behavioural data makes it difficult to discern scavenging from predation among extinct forms where it was surely exhibited to a similar extent among prehistoric vertebrate carnivores.
Some simple heuristics can inform us whether or not the prey was scavenged, for instance, in cases where it was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.}
But clearly, a scavenger does not only feed on animals too big for it to have hunted.
\textcolor{blue}{There are other methods however, that allow us to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper e.g. energetics models, comparative anatomy, palaeontology etc. \citep{shipman1986scavenging,ruxton2003could}. 
Yet, a synthesis describing the natural history of scavengers is absent from the literature.
Here we rectify this shortcoming in charting the natural history of scavenging by assessing its potential among extinct vertebrate groups given their ecology, functional traits and the environmental context.}

@@

\textcolor{blue}{Species that rely on scavenging to sustain substantial portions of their diets must encounter a sufficient amount of carrion in order to meet their energetic demands.
Once found, the scavenger must be able to out-compete any potential competitors and efficiently process the, typically decaying, carcass replete with micro-organism derived toxins \citep{ruxton2014fruit}.}
These characteristics can be assumed to be under evolutionary selection pressures for traits that increase carrion discovery and monopoly.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics \textcolor{blue}{\citep{moleon2014inter}. 
Even these aspects can be indirectly impacted by the behaviour of the scavenger \citep{moleon2014inter}.}
Each of these facets are essentially the backbone of fundamental ecological theory and are the key parameters defined in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}.  
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place \textcolor{blue}{an extinct vertebrate carnivore} and assess the likely importance of carrion in its diet.

@@

\section{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all determine the encounter rate between a scavenger and the carcasses it is searching for and we would thus expect selection pressures to act on the various traits that govern these parameters.
Selection might also drive a species to reduce its metabolic requirements so that it can survive long periods between meals, or expand a species ability to process tissues or decaying flesh that might ordinarily be discarded.
However, as we noted above, encounter rate is also determined by the productivity of the environment, and ultimately the rate at which carcasses are produced. 
\textcolor{blue}{Scavengers can drive this process indirectly, for instance, by consuming the majority of carrion, obligate scavengers deny a resource to their facultative counterparts obliging the latter to hunt more and thus produce more carrion \citep{moleon2014inter}.}

\subsection{Metabolism}
Because of the sporadic nature of carrion we would expect adaptations that reduce energetic costs of maintenance to be selected for in scavengers as it would allow for longer inter-feeding periods. 
Extant reptiles possess an advantage here, in that over the course of a year their food requirements can be 30 times lower than an endotherm of equal size \citep{Nagy1621}.
\cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they ``exhibit exceedingly low maintenance metabolisms, and most can survive on a few scant feedings per year.
It is, therefore, possible for snakes to rely largely on infrequent, less energy-rich meals."
In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is ``far more common than currently acknowledged."\citep{devault2002scavenging}.
The same reasoning can be applied to crocodiles and their allies \citep{forrest2003evidence,moleon2015carcass}. 
\cite{carey1982temperature} found that sharks, as ectothermics, have the ability to go weeks between meals because they focus on the energy-rich sections of carcasses (see Handling Time \ref{handlingTime}). 
Endotherms have also evolved physiological mechanisms that allow them to depress their otherwise high metabolic rates at certain times e.g. vultures who do so while resting at their roost and during periods of food deprivation \citep{bahat1998nocturnal}.

\subsection{Locomotion}
The \textcolor{blue}{relative} unpredictability of carrion also means scavenging depends more on the ability to efficiently move over larger areas than does predation.
This generally requires an efficient transfer of metabolic energy into movement which relies on the animal's anatomy and physiology as well as the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is found in flying species. 
Despite the energetic costs of flight, the only known vertebrate obligate scavengers are the old and the new world vultures. 
While powered flight is energetically expensive, species like vultures have evolved behavioural and anatomical features to exploit air currents using their large wingspans, allowing them to soar at a cost of only twice their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}.
By depending on thermal air flows these species can forage over vast ranges \citep{spiegel2013factors}. 
An analogous mode of locomotion is also exploited by seabirds, who use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid carcases, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

@@

Existing in this environment also means being able to gather a wealth of information from other foragers, be they conspecifics or otherwise \textcolor{blue}{\citep{jackson2008effect,KaneVul,moleon2014inter}}.
These interactions are properly three dimensional in the sense of \cite{pawar2012dimensionality} because both producer and scrounger are in the air \citep{dall2005information}. 
Again, returning to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}.
These behaviours allow them to forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \textcolor{blue}{\citep{cortes2014bird}}.
Information transfer of this kind is typically inadvertent and as a consequence no complex social interactions are required, simply the ability to recognise a successful forager.
\textcolor{blue}{Thus, given pterosaurs seem to have cohabited in large numbers \citep{witton2013pterosaurs}, and the known benefits this can have for social foraging in birds \citep{jackson2011evolutionary}, it seems probable that scrounging behaviours were seen in the flying pterosaurs as well.}

@@

\subsubsection{\textcolor{blue}{Biotic effects}}
Through facilitation e.g. local enhancement, adept scavengers may increase the amount of carrion mammalian carnivores feed on thus increasing their population size \citep{moleon2014inter}. 
This higher population of predators will take more prey and so produce more carcasses.
Conversely, by feeding on carcasses, predators may hunt less because they are sated by carrion, ultimately reducing predation risk on their prey base \citep{moleon2014inter}.

\textcolor{blue}{Modern scavenging assemblages can be influenced by carcass size \citep{moleon2015carcass}. 
Larger carcasses tend to last longer and also present a more conspicuous target for a foraging scavenger which results in more species attending them \citep{moleon2015carcass}.
This will have had implications for extinct assemblages because body size distributions vary across different environments but also across time. 
\cite{10.1371/journal.pone.0051925} showed Mesozoic faunal distributions may have been skewed towards larger species even when fossil biases are taken into account. 
If this pattern is real then the scavenger assemblages during this era would have been particularly diverse.}

@@

\textcolor{blue}{\subsection{Facilitation}
As well as competitive interactions, there are also facilitatory processes at play between vertebrate scavengers that can reduce handling time for the facilitated species. 
Rather than there being a random assortment of species at a carcass, scavenging assemblages tend to be nested \citep{selva2007nested}. 
This means the species that feed on the majority of carcasses, for example vultures, are a subset of a larger community which comprises hyenas, jackals, raptors etc. \citep{sebastian2016nested}. 
In extant systems these interactions can reduce handling times for the facilitated species. 
Some examples include vultures, hyenas and wolves opening the tough hides of ungulate carcasses that would otherwise be inaccessible to corvids and smaller mammalian carnivores \citep{selva2003scavenging,moleon2015carcass}.
The availability of carcasses can be indirectly influenced by the action of scavengers. 
Through facilitation e.g. local enhancement, vultures may increase the amount of carrion mammalian carnivores feed on thus increasing their population size \citep{moleon2014inter}. 
This higher population of predators will take more prey and so produce more carcasses. Conversely, by feeding on carcasses, predators may hunt less because they are sated by carrion, ultimately reducing predation risk on their prey base \citep{moleon2014inter}.}

\subsection{Food Processing}
Aside from coping with competitors, another vital component of carrion handling time is the ability to maximise the energy gain from the remains while reducing the energetics of doing so. 
At whale carcases, white and blue sharks are known to preferentially feed on the blubber layer \citep{long1996sharks}. 
Blubber is an energy rich portion of the carcass that can allow a shark to survive for 1.5 months on 30 kg of the material \citep{carey1982temperature}. 
On land many scavengers utilize late-stage carcass material that is less subject to decomposition and may be unavailable to other competitors, for example bone.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
\textcolor{blue}{Within mammals,} this ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), \textcolor{blue}{because} the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers (\citealt{palmqvist2011giant}; \textcolor{blue}{note, however that the inability of these cats to deal with bone may be overstated; \citealt{binder2010comparison}}). 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

>>>>

\begin{abstract} 
Despite its prevalence, scavenging is a difficult behaviour to observe in modern day carnivores and impossible to study directly in extinct species. 
Yet, there are certain intrinsic and environmental features of a species that push it towards a scavenging lifestyle. 
These can be thought of as some of the principal parameters in optimal foraging theory namely, encounter rate, handling time and prey availability. 
We use these components to highlight the morphologies and environments that would have been conducive to scavenging over geological time by focusing on the dominant vertebrate groups of the land, sea and air. 
The result is a synthesis on the natural history of scavenging. 
These features that make up our scale of scavenging can be applied to any given species and allow us to judge the likely importance of this foraging behaviour. 
\end{abstract}

@@

Historically, \textcolor{blue}{food webs have classified most acts of carnivory as predation events \citep{wilson2011scavenging}. 
This is clearly a mistaken view because of the existence of scavenging, a behaviour displayed by almost all carnivorous vertebrates \citep{devault2003scavenging}. 
Recent research has begun to redress this imbalance by recognising the prevalence of scavenging and detailing its ecological consequences across modern ecosystems such that are beginning to appreciate the implications for trophic ecology \citep{pereira2014facultative,periquet2015lion}.
Indeed, there is no discrete divide between predators and scavengers but rather a continuous gradient in terms of the proportion of carrion in the diet \citep{pereira2014facultative}.
Even vultures, the canonical example of obligate scavengers, can hunt \citep{margalida2011scavenger}. 
Unfortunately, the lack of direct behavioural data makes it difficult to discern scavenging from predation among extinct forms where it was surely exhibited by prehistoric vertebrate carnivores.
Some simple heuristics can inform us whether or not the prey was scavenged, for instance, in cases where it was simply too large to have been killed by the ostensible predator \citep{pobiner2008paleoecological}.}
But clearly, a scavenger does not only feed on animals too big for it to have hunted.

\textcolor{blue}{There are other methods however, that can allow us to discern the most suitable morphologies, physiologies and environments for a scavenging lifestyle to prosper e.g. energetics models, comparative anatomy, palaeontology etc. \citep{shipman1986scavenging,ruxton2003could}. 
Yet, a synthesis describing the natural history of scavengers is absent from the literature.
Here we rectify this shortcoming in charting the natural history of scavenging by assessing its potential among vertebrate groups past and present given their ecology, functional traits and the environmental context.}

@@

\textcolor{blue}{Species that rely on scavenging to sustain substantial portions of their diets must encounter a sufficient amount of carrion in order to meet their energetic demands.
Once found, the scavenger must be able to out-compete any potential competitors and efficiently process the, typically decaying, carcass replete with micro-organism derived toxins \citep{ruxton2014fruit}.}
These characteristics can be assumed to be under evolutionary selection pressures for traits that increase carrion discovery and monopoly.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics \textcolor{blue}{\citep{moleon2014inter}. 
Even these aspects can be indirectly impacted by the behaviour of the scavenger \citep{moleon2014inter}.}
Each of these facets are essentially the backbone of fundamental ecological theory and are the key parameters defined in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}.  
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place \textcolor{blue}{a given vertebrate carnivore} and assess the likely importance of carrion in its diet.

@@

\section{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all determine the encounter rate between a scavenger and the carcasses it is searching for and we would thus expect selection pressures to act on the various traits that govern these parameters.
Selection might also drive a species to reduce its metabolic requirements so that it can survive long periods between meals, or expand a species' ability to process tissues or decaying flesh that might ordinarily be discarded.
However, as we noted above, encounter rate is also determined by the productivity of the environment, and ultimately the rate at which carcasses are produced. 
\textcolor{blue}{Scavengers can drive this process indirectly, for instance, by consuming the majority of carrion, more obligate scavengers deny a resource to their facultative counterparts obliging the latter to hunt more, and thus, increase the amount of carrion \citep{moleon2014inter}.}

\subsection{Metabolism}
Because of the ephemeral nature of carrion \citep{devault2003scavenging,ruxton2004obligate} we would expect adaptations that reduce energetic costs of maintenance to be selected for in scavengers as it would allow for longer inter-feeding periods. 
Extant reptiles possess an advantage here, in that over the course of a year their food requirements can be 30 times lower than an endotherm of equal size \citep{Nagy1621}.
\cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they ``exhibit exceedingly low maintenance metabolisms, and most can survive on a few scant feedings per year.
It is, therefore, possible for snakes to rely largely on infrequent, less energy-rich meals."
In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is ``far more common than currently acknowledged."\citep{devault2002scavenging}.
The same reasoning can be applied to crocodiles and their allies \citep{forrest2003evidence,moleon2015carcass}. 
\cite{carey1982temperature} found that sharks, as ectothermics, have the ability to go weeks between meals because they focus on the energy-rich sections of carcasses (see Handling Time \ref{handlingTime}). 
Endotherms have also evolved physiological mechanisms that allow them to depress their otherwise high metabolic rates at certain times e.g. vultures who do so while resting at their roost and during periods of food deprivation \citep{bahat1998nocturnal}.

\subsection{Locomotion}
Carrion can often be unpredictable, such as in cases where animals succumb to disease \citep{devault2003scavenging}, which means scavenging depends more on the ability to efficiently move over larger areas than does predation.
This generally requires an efficient transfer of metabolic energy into movement which relies on the animal's anatomy and physiology as well as the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is found in flying species. 
Despite the energetic costs of flight, the only known vertebrates that can be obligate scavengers are the old and the new world vultures. 
While powered flight is energetically expensive, species like vultures have evolved behavioural and anatomical features to exploit air currents using their large wingspans, allowing them to soar at a cost of only around 1.5 times their metabolic rate \citep{hedenstrom1993migration,duriez2014cheap}.
By depending on thermal air flows these species can forage over vast ranges \citep{spiegel2013factors}. 
An analogous mode of locomotion is also exploited by seabirds, who use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid carcases, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

@@

@@

\subsubsection{\textcolor{blue}{Biotic effects}}
Foragers do not exist in isolation and we know from field observations that scavengers can scrounge on the discoveries of other carnivores. 
This sort of facilitation occurs in both obligate and facultative scavengers, in the air and on the ground \citep{KaneVul,jones2015african}.

In flight, birds are able to gather a wealth of information from other foragers, be they conspecifics or otherwise \textcolor{blue}{\citep{jackson2008effect,KaneVul,moleon2014inter}}.
% These interactions are properly three dimensional in the sense of \cite{pawar2012dimensionality} because both producer and scrounger are in the air \citep{dall2005information}. 
Again, returning to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}.
These behaviours allow them to forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \textcolor{blue}{\citep{cortes2014bird}}.
Information transfer of this kind is typically inadvertent and as a consequence no complex social interactions are required, simply the ability to recognise a successful forager.
\textcolor{blue}{Thus, given pterosaurs seem to have cohabited in large numbers \citep{witton2013pterosaurs}, and the known benefits this can have for social foraging in birds \citep{jackson2011evolutionary}, it seems probable that scrounging behaviours were seen in the flying pterosaurs as well.}

This type of facilitation then increases the encounter rate of the facilitated species and can increase the population size of the latter \citep{moleon2014inter}. 
This higher population of predators may take more prey and so produce more carcasses.
Conversely, by feeding on carcasses, predators may hunt less because they are sated by carrion, ultimately reducing predation risk on their prey base \citep{moleon2014inter}.

\textcolor{blue}{Modern scavenging assemblages are known to be influenced by carcass size \citep{moleon2015carcass}. 
Larger carcasses tend to last longer and also present a more conspicuous target for a foraging scavenger which results in more species attending them \citep{moleon2015carcass}.
This will have had implications for extinct assemblages because body size distributions vary across different environments but also across time. 
\cite{10.1371/journal.pone.0051925} showed Mesozoic faunal distributions may have been skewed towards larger species even when fossil biases are taken into account. 
Similarly, the megafauna of the Pleistocene \citep{doughty2013legacy} would have produced large carcasses. 
As a result, the scavenger assemblages during this era would have been particularly diverse.}

@@

\textcolor{blue}{\subsection{Facilitation}
In contrast to competitive interactions, there are facilitatory processes at play between vertebrate scavengers that can benefit the facilitated species. 
Rather than there being a random assortment of species at a carcass, scavenging assemblages tend to be nested \citep{selva2007nested}. 
This means the species that feed on the majority of carcasses, for example vultures, are a subset of a larger community which comprises hyenas, jackals, raptors etc. \citep{sebastian2016nested}. 
We noted above that this often results in an increased encounter rate but it can also reduce handling time for the species that follow. 
% In extant systems these interactions can reduce handling times for the facilitated species. 
Some examples of reduced handling time include vultures, hyenas and wolves opening the tough hides of ungulate carcasses that would otherwise be inaccessible to corvids and smaller mammalian carnivores \citep{selva2003scavenging,moleon2015carcass}.}

\subsection{Food Processing}
Another vital component of carrion handling time is the ability to maximise the energy gain from the remains while reducing the energetics of doing so. 
At whale carcases, white and blue sharks are known to preferentially feed on the blubber layer \citep{long1996sharks}. 
Blubber is an energy rich portion of the carcass that can allow a shark to survive for 1.5 months on 30 kg of the material \citep{carey1982temperature}. 
On land many scavengers utilize late-stage carcass material that is less subject to decomposition and may be unavailable to other competitors, for example bone.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
\textcolor{blue}{Within mammals,} this ability reached its zenith among hyenas with the evolution of the estimated 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), \textcolor{blue}{because} the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers (\citealt{palmqvist2011giant}; \textcolor{blue}{note, however that the inability of these cats to deal with bone may be overstated; \citealt{binder2010comparison}}). 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.
